---
title: "Example trees"
author: "Martin R. Smith"
date: "22 November 2017"
output: rmarkdown::html_document
bibliography: example-trees.bib
vignette: >
  %\VignetteIndexEntry{Example Trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r Initialize, include=FALSE}
require('SlowQuartet')
set.seed(1) # 1 produces a random tree with a Qt dist of 226 ~= 330 * 2 / 3

# Colourblind palette from http://mkweb.bcgsc.ca/biovis2012/color-blindness-palette.png
tip_colours <- c(
  rgb(000,000,000, maxColorValue=255),
  rgb(000,073,073, maxColorValue=255),
  rgb(000,146,146, maxColorValue=255),
  rgb(255,109,182, maxColorValue=255),
  rgb(255,182,219, maxColorValue=255),
  rgb(073,000,146, maxColorValue=255),
  rgb(000,109,219, maxColorValue=255),
  rgb(182,109,255, maxColorValue=255),
  rgb(109,182,255, maxColorValue=255),
  rgb(182,219,255, maxColorValue=255),
  rgb(146,000,000, maxColorValue=255),
  rgb(146,073,000, maxColorValue=255),
  rgb(219,109,000, maxColorValue=255),
  rgb(036,255,036, maxColorValue=255),
  rgb(255,255,109, maxColorValue=255)
)[-c(4, 7)] # Rm Tritamopia duplicates of 13 and 3
```

# Tree distance metrics

The advantages of the Quartet distance [@Estabrook1985] over other tree distance metrics are
best illustrated by examining a contrived set of example trees.

Here are some example trees.  I've made a reference tree, then peturbed it by
some simple transformations: moving one taxon (labelled 1) or a pair of taxa
(10, 11) a short, medium or large distance from their original position.

In certain trees, I have also collapsed particular nodes.

`opposite_tree` is a tree designed to be as different as possible from the 
reference tree. `random_tree` is generated at random.

```{r Example trees}
n_tip <- 11
ref_tree <- ape::read.tree(text="(((1, 2), 3), (((4, 5), 6), ((7, (8, 9)), (10, 11))));")
trs <- test_trees <- list (
  ref_tree      = ref_tree,
  move_one_near = ape::read.tree(text="(((2, 3), 1), (((4, 5), 6), ((7, (8, 9)), (10, 11))));"),
  move_one_mid  = ape::read.tree(text="((2, 3), ((((1, 4), 5), 6), ((7, (8, 9)), (10, 11))));"),
  move_one_far  = ape::read.tree(text="((2, 3), (((4, 5), 6), ((7, (8, 9)), (10, (11, 1)))));"),
  move_two_near = ape::read.tree(text="(((1, 2), 3), (((4, 5), 6), ((7, (10, 11)), (8, 9))));"),
  move_two_mid  = ape::read.tree(text="(((1, 2), 3), ((((4, 5), (10, 11)), 6), (7, (8, 9))));"),
  move_two_far  = ape::read.tree(text="((((1, (10, 11)), 2), 3), (((4, 5), 6), (7, (8, 9))));"),
  collapse_one  = ape::read.tree(text="(((1, 2), 3), (((4, 5), 6), ((7, 8, 9), (10, 11))));"),
  collapse_some = ape::read.tree(text="((1, 2, 3, 4, 5, 6), ((7, 8, 9), (10, 11)));"),
  m1mid_col1    = ape::read.tree(text="((2, 3), ((((1, 4), 5), 6), ((7, 8, 9), (10, 11))));"),
  m1mid_colsome = ape::read.tree(text="((2, 3), ((((1, 4), 5), 6), (7, 8, 9, 10, 11)));"),
  m2mid_col1    = ape::read.tree(text="(((1, 2), 3), ((((4, 5), (10, 11)), 6), (7, 8, 9)));"),
  m2mid_colsome = ape::read.tree(text="(((1, 2), 3), (4, (10, 11), 5, 6, 7, 8, 9));"),
  opposite_tree = ape::read.tree(text="(((1, 11), 3), (((4, 9), 6), ((10, (8, 2)), (5, 7))));"),
  random_tree   = ape::rtree(n_tip, tip.label=seq_len(n_tip), br=NULL)
)
```

```{r Plot trees, echo=FALSE}
colplot <- function (tr, title=NULL, bold=NULL) {
  tr$edge.length <- rep(1, dim(tr$edge)[1])
  font <- rep(1, length(tr$tip.label))
  if (!is.null(bold)) font[tr$tip.label %in% bold] <- 4
  plot(tr, tip.col=tip_colours[as.integer(tr$tip.label)], main=title, cex.main=0.8, font=font)
}

par(mfrow=c(3,5), mar=rep(1.2, 4), cex.main=2/3, cex=0.9)
for (i in seq_along(test_trees)) colplot(test_trees[[i]], names(test_trees)[i])

```

## Moving a single taxon

If trees differ only in the location of a single taxon, then the distance between
two trees should correspond to the distance that this taxon has been moved.

```{R Moving a single taxon, fig.height=3, fig.width=12, echo=FALSE}
par(mfrow=c(1, 4), mar=rep(0.4, 4), cex=1)
colplot(ref_tree, 'Reference tree')
rfplot <- function (tr, title, highlight, spr=TRUE) {
  tree_dist <- phangorn::treedist(tr, ref_tree)
  tree_pair <- lapply(list(tr, ref_tree), ape::root, outgroup='1', resolve.root=FALSE)
  class(tree_pair) <- 'multiPhylo'
  topo_dist <- as.matrix(ape::dist.topo(tree_pair, method='PH85'))[2]
  
  colplot(tr, title, highlight)
  text(0, 10.6, paste0('Quartet = ', choose(11, 4) - MatchingQuartets(list(tr, ref_tree))[1, 2], '/', choose(11,4)), cex=0.8, pos=4)
  text(0,  9.6, paste0("RF = ", phangorn::RF.dist(tr, ref_tree)), cex=0.8, pos=4)
  text(0,  8.6, paste0("Symmetric = ", tree_dist[1]), cex=0.8, pos=4)
  text(0,  7.6, paste0("PH85 = ", topo_dist), cex=0.8, pos=4)
  text(0,  6.6, paste0("Path = ", signif(tree_dist[2], 3)), cex=0.8, pos=4)
  if (spr) text(0,  5.6, paste0("SPR = ", 1), cex=0.8, pos=4)
}
rfplot(trs$move_one_near, 'Short move', 1)
rfplot(trs$move_one_mid, 'Medium move', 1)
rfplot(trs$move_one_far,   'Long move', 1)

```

Note that all these trees have an SPR (or TBR) distance of 1.
Moreover, the RF distances for a medium or long move are equal, whereas the Quartet
distances identify the Long move as representing a greater modification of the
reference tree.

## Moving two taxa
Surprisingly, if two rather than one taxa are moved, RF distances recover a smaller 
distance:

```{R Moving a cherry, fig.height=8, fig.width=9, echo=FALSE}
par(mfrow=c(3, 3), mar=c(2.4, 0.4, 0.4, 0.4), cex=1)

colplot(tt$ref_tree, 'Reference tree')
rfplot(trs$move_one_near, 'Short move 1', 1)
rfplot(trs$move_two_near, 'Short move 2', 10:11)

colplot(tt$ref_tree, 'Reference tree')
rfplot(trs$move_one_mid, 'Medium move 1', 1)
rfplot(trs$move_two_mid, 'Medium move 2', 10:11)

colplot(tt$ref_tree, 'Reference tree')
rfplot(trs$move_one_far,   'Long move 1', 1)
rfplot(trs$move_two_far,   'Long move 2', 10:11)

```

## Unresolved trees

A further benefit is that the Quartet distance can meaningfully score trees
where not every node is resolved as bifurcating.


```{R Polytomies - same topology, fig.height=3, fig.width=9, echo=FALSE}
par(mfrow=c(1, 3), mar=c(2.4, 0.4, 0.4, 0.4), cex=1)

polyplot <- function (tr, title, highlight) {
  tree_pair <- lapply(list(tr, ref_tree), ape::root, outgroup='1', resolve.root=FALSE)
  class(tree_pair) <- 'multiPhylo'
  topo_dist <- as.matrix(ape::dist.topo(tree_pair, method='PH85'))[2]
  
  colplot(tr, title, highlight)
  text(0, 10.6, paste0('Quartets contradicted = ', choose(11, 4) - sum(MatchingQuartets(list(tr, ref_tree))[, 2]), '/', choose(11,4)), cex=0.8, pos=4)
  text(0,  9.6, paste0('Quartets unresolved = ', MatchingQuartets(list(tr, ref_tree))[2, 2], '/', choose(11,4)), cex=0.8, pos=4)
  text(0,  8.6, paste0("PH85 = ", topo_dist), cex=0.8, pos=4)
}

colplot(tt$ref_tree, 'Reference tree')
polyplot(trs$collapse_one, 'One polytomy', 7:9)
polyplot(trs$collapse_some, 'Much polytomy', 1:9)

```

In the above case, the underlying topology was unchanged.  We can also change
the topology -- here we have the same loss of resolution with the trees
in which one taxon (1) was moved:

```{R Polytomies, fig.height=3, fig.width=9, echo=FALSE}
par(mfrow=c(1, 3), mar=c(2.4, 0.4, 0.4, 0.4), cex=1)

polyplot(trs$move_one_mid, 'One moved, no polytomy', 1)
polyplot(trs$m1mid_col1, 'One moved, one polytomy', 1)
polyplot(trs$m1mid_colsome, 'One moved, much polytomy', 1)

```
And here again, collapsing nodes in the trees
in which two taxa (10 & 11) were moved:

```{R Polytomies 2, fig.height=3, fig.width=9, echo=FALSE}
par(mfrow=c(1, 3), mar=c(2.4, 0.4, 0.4, 0.4), cex=1)

polyplot(trs$move_two_mid,  'Two moved, no polytomy'  , 10:11)
polyplot(trs$m2mid_col1,    'Two moved, one polytomy' , 10:11)
polyplot(trs$m2mid_colsome, 'Two moved, much polytomy', 10:11)

```

## Maximum distance
Finally, a distance metric should distinguish slightly-peturbed trees from
random trees and those that are more different from the starting tree than 
expected by chance.

```{R Increasing distances, fig.height=3, fig.width=12, echo=FALSE}
par(mfrow=c(1, 4), mar=rep(0.4, 4), cex=1)
colplot(tt$ref_tree, 'Reference tree')
rfplot(trs$move_one_far,  'Small peturbation', 1, spr=FALSE)
rfplot(trs$random_tree,   'Random tree',       1, spr=FALSE)
rfplot(trs$opposite_tree, 'Maximum distance',  1, spr=FALSE)
```

## References

