---
title: "Example trees"
author: "Martin R. Smith"
date: "21 November 2017"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Example Trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r Initialize, include=FALSE}
require('SlowQuartet')
set.seed(100)

# Colourblind palette from http://mkweb.bcgsc.ca/biovis2012/color-blindness-palette.png
tip_colours <- c(
  rgb(000,000,000, maxColorValue=255),
  rgb(000,073,073, maxColorValue=255),
  rgb(000,146,146, maxColorValue=255),
  rgb(255,109,182, maxColorValue=255),
  rgb(255,182,219, maxColorValue=255),
  rgb(073,000,146, maxColorValue=255),
  rgb(000,109,219, maxColorValue=255),
  rgb(182,109,255, maxColorValue=255),
  rgb(109,182,255, maxColorValue=255),
  rgb(182,219,255, maxColorValue=255),
  rgb(146,000,000, maxColorValue=255),
  rgb(146,073,000, maxColorValue=255),
  rgb(219,109,000, maxColorValue=255),
  rgb(036,255,036, maxColorValue=255),
  rgb(255,255,109, maxColorValue=255)
)[-c(4, 7)] # Rm Tritamopia duplicates of 13 and 3
```

# Tree distance metrics

The advantages of the Quartet distance over other tree distance metrics are
best illustrated by examining a contrived set of example trees.

Here are some example trees.  I've made a reference tree, then peturbed it by
some simple transformations: moving one taxon (labelled 1) or a pair of taxa
(10, 11) a short, medium or large distance from their original position.

In certain trees, I have also collapsed particular nodes.

`opposite_tree` is a tree designed to be as different as possible from the 
reference tree. `random_tree` is generated at random.

```{r Example trees}
n_tip <- 11
ref_tree <- ape::read.tree(text="(((1, 2), 3), (((4, 5), 6), ((7, (8, 9)), (10, 11))));")
trs <- test_trees <- list (
  ref_tree      = ref_tree,
  move_one_near = ape::read.tree(text="(((2, 3), 1), (((4, 5), 6), ((7, (8, 9)), (10, 11))));"),
  move_one_mid  = ape::read.tree(text="((2, 3), ((((1, 4), 5), 6), ((7, (8, 9)), (10, 11))));"),
  move_one_far  = ape::read.tree(text="((2, 3), (((4, 5), 6), ((7, (8, 9)), (10, (11, 1)))));"),
  move_two_near = ape::read.tree(text="(((1, 2), 3), (((4, 5), 6), ((7, (10, 11)), (8, 9))));"),
  move_two_mid  = ape::read.tree(text="(((1, 2), 3), ((((4, (10, 11)), 5), 6), (7, (8, 9))));"),
  move_two_far  = ape::read.tree(text="((((1, (10, 11)), 2), 3), (((4, 5), 6), (7, (8, 9))));"),
  collapse_one  = ape::read.tree(text="(((1, 2), 3), (((4, 5), 6), ((7, 8, 9), (10, 11))));"),
  collapse_some = ape::read.tree(text="((1, 2, 3, 4, 5, 6), ((7, 8, 9), (10, 11)));"),
  m1mid_col1    = ape::read.tree(text="((2, 3), ((((1, 4), 5), 6), ((7, 8, 9), (10, 11))));"),
  m1mid_colsome = ape::read.tree(text="((2, 3), ((((1, 4), 5), 6), (7, 8, 9, 10, 11)));"),
  m2mid_col1    = ape::read.tree(text="(((1, 2), 3), ((((4, (10, 11)), 5), 6), (7, 8, 9)));"),
  m2mid_colsome = ape::read.tree(text="(((1, 2), 3), ((4, (10, 11)), 5, 6, 7, 8, 9));"),
  opposite_tree = ape::read.tree(text="(((1, 11), 3), (((4, 9), 6), ((10, (8, 2)), (5, 7))));"),
  random_tree   = ape::rtree(n_tip, tip.label=seq_len(n_tip), br=NULL)
)
```

```{r Plot trees, echo=FALSE}
colplot <- function (tr, title=NULL, bold=NULL) {
  tr$edge.length <- rep(1, dim(tr$edge)[1])
  font <- rep(1, length(tr$tip.label))
  if (!is.null(bold)) font[tr$tip.label %in% bold] <- 4
  plot(tr, tip.col=tip_colours[as.integer(tr$tip.label)], main=title, cex.main=0.8, font=font)
}

par(mfrow=c(3,5), mar=rep(1.2, 4), cex.main=2/3, cex=0.9)
for (i in seq_along(test_trees)) colplot(test_trees[[i]], names(test_trees)[i])

```

## Moving a single taxon

If trees differ only in the location of a single taxon, then the distance between
two trees should correspond to the distance that this taxon has been moved.

```{R Moving a single taxon, fig.height=3, fig.width=12, echo=FALSE}
par(mfrow=c(1, 4), mar=rep(0.4, 4), cex=1)
colplot(tt$ref_tree, 'Reference tree')
rfplot <- function (tr, title, highlight) {
  colplot(tr, title, highlight)
  text(0, 10.5, paste0('Quartet = ', choose(11, 4) - MatchingQuartets(list(tr, ref_tree))[1, 2], '/', choose(11,4)), cex=0.8, pos=4)
  text(0,  9.5, paste0("RF = ", phangorn::RF.dist(tr, ref_tree)), cex=0.8, pos=4)
  text(0,  8.5, paste0("SPR = ", 1), cex=0.8, pos=4)
}
rfplot(trs$move_one_near, 'Short move', 1)
rfplot(trs$move_one_mid, 'Medium move', 1)
rfplot(trs$move_one_far,   'Long move', 1)

```

Note that all these trees have an SPR (or TBR) distance of 1.
Moreover, the RF distances for a medium or long move are equal, whereas the Quartet
distances identify the Long move as representing a greater modification of the
reference tree.

Surprisingly, if two rather than one taxa are moved, RF distances recover a smaller 
distance:

```{R Moving a cherry, fig.height=6, fig.width=9, echo=FALSE}
par(mfrow=c(3, 3), mar=c(2.4, 0.4, 0.4, 0.4), cex=1)

colplot(tt$ref_tree, 'Reference tree')
rfplot(trs$move_one_near, 'Short move 1', 1)
rfplot(trs$move_two_near, 'Short move 2', 10:11)

colplot(tt$ref_tree, 'Reference tree')
rfplot(trs$move_one_mid, 'Medium move 1', 1)
rfplot(trs$move_two_mid, 'Medium move 2', 10:11)

colplot(tt$ref_tree, 'Reference tree')
rfplot(trs$move_one_far,   'Long move 1', 1)
rfplot(trs$move_two_far,   'Long move 2', 10:11)

```


```{R Calculate distances, message=FALSE, warning=FALSE}

dist_measures <- c('symmetric', 'path', 'R-F', 'uSPR', 'TBR', 'PH85', '4tet_=', '4tet_?', '4tet_X')
dists <- matrix(NA, nrow=length(dist_measures), ncol=length(test_trees))
rownames(dists) <- dist_measures
colnames(dists) <- names(test_trees)
dists[1:2, ] <- vapply(test_trees, phangorn::treedist, tree2=ref_tree, double(2))
dists[3, ] <- vapply(test_trees, phangorn::RF.dist, tree2=ref_tree, double(1))
# Exact distance not calculated for latter trees, but will be highish
dists[4, ] <- c(0, 1, 1, 1, 1, 1, 1, NA, NA, NA, NA, NA, NA, 10, 10)
dists[5, ] <- c(0, 1, 1, 1, 1, 1, 1, NA, NA, NA, NA, NA, NA, 10, 10) 

class(test_trees) <- 'multiPhylo'
dists[6, ] <- as.matrix(ape::dist.topo(test_trees, method='PH85'))[, 1]

n_quartets <- choose(n_tip, 4)
quartet_matches <- MatchingQuartets(test_trees)
other_trees <- seq_along(test_trees)[-1]

dists[7:8, ] <- quartet_matches / n_quartets * 100
dists[9, ] <- (n_quartets - colSums(quartet_matches)) / n_quartets * 100
dists[7:9, 1] <- c(100, 0, 0)
dists

```
