---
title: "Getting started with Quartet"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
  bookdown::pdf_document2:
    includes:
      in_header: ../inst/preamble.tex
    number_sections: false
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa-old-doi-prefix.csl
vignette: >
  %\VignetteIndexEntry{Getting started with Quartet}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

This document should contain all you need to get started measuring tree 
distances with 'Quartet'.  If you get stuck, please 
[let me know](https://github.com/ms609/Quartet/issues/new?title=Suggestion:+)
so I can improve this documentation.

## Loading trees

Instructions for loading phylogenetic trees into R can be found in a [separate vignette](https://ms609.github.io/TreeTools/articles/load-trees.html).
For these examples, we'll enter two simple trees by hand:

```{r set-up}
tree1 <- ape::read.tree(text = '(A, ((B, (C, (D, E))), ((F, G), (H, I))));')
tree2 <- ape::read.tree(text = '(A, ((B, (C, (D, (H, I)))), ((F, G), E)));')
```

## Calculating distances

We can calculate distances between pairs of trees using the 'Quartet' package.

First we'll install the package.
We can either install the stable version from the CRAN repository:

```r
install.packages('Quartet')
```

or the development version, from GitHub -- which will contain the latest
features but may not be as extensively tested:

```r 
devtools::install_github('ms609/Quartet')
```

Then we'll load the package into R's working environment:

```{r load-package, message=FALSE}
library('Quartet')
```

Now the package's functions are available within R.
Let's proceed to calculate some tree distances.

### Pairs of trees

Calculating the distance between two trees is a two stage process.
For a quartet distance, we first have to calculate the status of each quartet:

```{r quartet-status}
statuses <- QuartetStatus(tree1, tree2)
```

Then we convert these counts into a distance metric (or similarity measure)
that suits our needs -- perhaps the Quartet Divergence:

```{r measure-distance}
QuartetDivergence(statuses, similarity = FALSE)
```

We can calculate all similarity metrics at once using:

```{r all-metrics}
SimilarityMetrics(statuses, similarity = TRUE)
```

It can be instructive to visualize how each split in the tree is contributing
to the quartet similarity:

```{r visualize-quartets}
VisualizeQuartets(tree1, tree2)
```

Rather than using quartets, we might want to use partitions as the basis of our
comparison:

```{r partitions}
SimilarityMetrics(SplitStatus(tree1, tree2))
```


### Multiple comparisons

If you have more than two trees to compare, you can send a list of trees 
(class: `list` or `multiPhylo`) to the distance comparison function.

You can calculate the similarity between one tree and a forest of other trees:

```{r multi-trees}
library('TreeTools', quietly = TRUE, warn.conflicts = FALSE)
oneTree <- CollapseNode(as.phylo(0, 11), 14)
twoTrees <- structure(list(bal = BalancedTree(11), pec = PectinateTree(11)),
                      class = 'multiPhylo')

status <- SharedQuartetStatus(twoTrees, cf = oneTree)
QuartetDivergence(status)
```

Or between one tree and (itself and) all other trees in the forest:
```{r one-to-many}
forest <- as.phylo(0:5, 11)
names(forest) <- letters[1:6]
status <- SharedQuartetStatus(forest)
QuartetDivergence(status)
```

Or between each pair of trees in a forest:
```{r many-to-many}
status <- ManyToManyQuartetAgreement(forest)
QuartetDivergence(status, similarity = FALSE)
```

Or between one list of trees and a second:
```{r pairwise}
status <- TwoListQuartetAgreement(forest[1:4], forest[5:6])
QuartetDivergence(status, similarity = FALSE)
```

## Trees with different tip labels

"Quartet" can compare trees of different sizes or with non-identical sets of
taxa.  Quartets pertaining to a leaf that does not occur in one tree are treated
as unresolved.

```{r different-tips, fig.height = 1.5, fig.width = 4, fig.align = "center"}
treeAG <- PectinateTree(letters[1:7])
treeBI <- PectinateTree(letters[2:9])
treeEJ <- PectinateTree(letters[5:10])
par(mfrow = c(1, 3), mar = rep(0.3, 4), cex = 1)
plot(treeAG); plot(treeBI); plot(treeEJ)

QuartetState(letters[1:4], treeAG) # 3: C is closest to D
QuartetState(letters[1:4], treeBI) # 0: unresolved in this tree

# Calculate status for all leaves observed in trees: here, A..I
QuartetStatus(treeAG, treeBI, nTip = TRUE)

# Calculate status for specified number of leaves
# Here, we have ten taxa A..J, but J does not occur in either of these trees
QuartetStatus(treeAG, treeBI, nTip = 10)

# Compare a list of trees with different numbers of leaves to a reference
QuartetStatus(c(treeAG, treeBI, treeEJ), cf = treeAG, nTip = TRUE)

# Compare all pairs of trees in a list.
# "u" shows how many possible quartets are unresolved in both trees
ManyToManyQuartetAgreement(c(treeAG, treeBI, treeEJ), nTip = TRUE)[, , "u"]
```


## Other calculations

To calculate how many quartets are unique to a certain tree (akin to
the partitionwise equivalent `ape::prop.clades`), use:
```{r in-one-only}
interestingTree <- as.phylo(42, 7)
referenceTrees <- list(BalancedTree(7), PectinateTree(7))
status <- CompareQuartetsMulti(interestingTree, referenceTrees)
```

`status['x_only']` = `r status['x_only']` quartets are resolved in a certain way
in `interestingTree`, but not resolved that way in any `referenceTrees`.

## What next?

You may wish to:

- [Read more](https://ms609.github.io/Quartet/articles/Quartet-Distance.html) about Quartet distances

- Review [alternative distance measures](https://ms609.github.io/TreeDist/index.html)
  and corresponding [functions](https://ms609.github.io/TreeDist/reference/index.html#section-tree-distance-measures)

- [Interpret or contextualize](https://ms609.github.io/TreeDist/articles/using-distances.html) 
  tree distance metrics
